---
title: "P8105 Data Science I - Homework 2"
author: "Kevin S.Wongsodirdjo   UNI: ksw2137"
date: "9/24/2019"
output: github_document
---

```{r setup, include = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE, warning = FALSE)
```

# Introduction
R Markdown document for P8105 homework 2. Code chunk below shows that we are loading the required packages in order to manipulate the imported data.

```{r package_load, message = FALSE, warning = FALSE}

# Loads the necessary packages
Packages <- c("tidyverse", "readxl")
invisible(lapply(Packages, library, character.only = TRUE))

```


# Problem 1 - Data Import for Mr. Trash Wheel, 2017 Precipitation, & 2018 Precipitation and Data Clean-up
We are loading datasets from the Healthy Harbor Water Wheel dataset, particularly from the sheets: "Mr. Trash Wheel", "2018 Precipitation", and "2017 Precipitation". While the Mr. Trash Wheel dataset is generally "tidy", we will modify these further to make it easier to work with the data. Dataset for precipitation from 2017 and 2018 were on separate sheets and needed to be merged since we would like a combined dataset. 

```{r import_trash_wheel}

# Import excel sheet and piping to clean data for Mr. Trash Wheel
trash_wheel_data <- 
  read_excel("./data/trash_wheel/HealthyHarborWaterWheelTotals2018-7-28.xlsx",
             sheet = "Mr. Trash Wheel") %>%                         # Specifies which sheet to read
  janitor::clean_names() %>%                                        # renames column titles as x_y and lowercases
  select(-x15) %>%                                                  # removes the "x15"(comments)
  drop_na(year) %>%                                                 # drops any rows containing NA in "year" column
  mutate(
    date = format(date, format = "%d"),                             # Changes the date format from %y%m%d to %d
    sports_balls = round(sports_balls, 0),                          # Rounds the numbers in sports_balls to integers
    sports_balls = as.integer(sports_balls),                        # Converts type from double to integers
    month = str_to_lower(month)) %>%                                # Makes months to have lowercases
  select(dumpster, year, month, everything())                       # Re-arranges the column to be "tidier"

# separates sportsball column to be used somewhere else
col_sportball <- pull(trash_wheel_data, sports_balls)




# Reads the 2017 precipitation data
precip2017_data <- 
  read_excel("./data/trash_wheel/HealthyHarborWaterWheelTotals2018-7-28.xlsx", 
             sheet = "2017 Precipitation",                          # Specifies the sheet
             range = "A2:B14") %>%                                  # Specifies range of cells to be read
  janitor::clean_names() %>%                                        # turns column names to lowercases
  mutate(year = "2017") %>%                                         # Adds year column
  select(year, everything()) %>%                                    # Re-arrange so year is 1st column
  drop_na(total)                                                    # drops any blank cells in "total" column
  


# Reads the 2018 precipitation data    
precip2018_data <- 
  read_excel("./data/trash_wheel/HealthyHarborWaterWheelTotals2018-7-28.xlsx", 
             sheet = "2018 Precipitation",                          # Specifies the sheet
             range = "A2:B14") %>%                                  # Specifies range of cells to be read
  janitor::clean_names() %>%                                        # turns column names to lowercases
  mutate(year = "2018") %>%                                         # Adds year column
  select(year, everything()) %>%                                    # Re-arrange so year is 1st column 
  drop_na(total)                                                    # drops any blank cells in "total" column


# New dataframe that merged precip2017 and precip2018 by year, then month, then "total". 
merged_precip20178_data <- full_join(precip2017_data, precip2018_data, by = c("year", "month", "total")) %>% 
  mutate(month = str_to_lower(month.name[month])) %>%               # changes month from "1" to "january" and so on
  rename(monthly_precip = total)                                    # renames the total column to monthly_precip

# Pulls the monthly precipitate column for other uses
total_precip2018 <- sum(pull(precip2018_data, total))

```

Data from Mr. Trash Wheel was obtained from Healthy Harbor Water Wheel's website and contains a total of `r ncol(trash_wheel_data)` variables containing `r nrow(trash_wheel_data)` observations each. Examples of particularly key variables included are `volume_cubic_yards` (trash volume in cubic yards) and unique types of trash such as `cigarette_butts` or `plastic_bottles`. Interestingly, there were also `sports_balls` included, which has a total median of `r median(col_sportball)` balls per dumpster. 

The precipitation data was less interesting; there were a total of `r nrow(merged_precip20178_data)` observations corresponding to each month and its precipitate, which is under the `monthly_precip` column. The `year` identifier was added and merged as it was not included in the original (already separated by sheets). Unfortunately, the 2018 precipitation data was only up to July, thus the `r nrow(precip2018_data)` observations; consequently, total measured precipitation for 2018 was `r total_precip2018`.

# Problem 2

This problem uses the FiveThirtyEight data; these data were gathered to create the interactive graphic on this page. In particular, we’ll use the data in pols-month.csv, unemployment.csv, and snp.csv. Our goal is to merge these into a single data frame using year and month as keys across datasets.

First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove  prez_dem and prez_gop; and remove the day variable.

Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

Join the datasets by merging snp into pols, and merging unemployment into the result.

Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

Note: we could have used a date variable as a key instead of creating year and month keys; doing so would help with some kinds of plotting, and be a more accurate representation of the data. Date formats are tricky, though. For more information check out the lubridate package in the tidyverse.

# Problem 3

This problem uses data from NYC Open data on the popularity of baby names, and can be downloaded here.

Load and tidy the data. Note that, although these data may seem fairly well formatted initially, the names of a categorical predictor and the case structure of string variables changed over time; you’ll need to address this in your data cleaning. Also, some rows seem duplicated, and these will need to be removed (hint: google something like “dplyr remove duplicate rows” to get started).

Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year. Produce a similar table showing the most popular name among male children over time.

Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis).

